# Data Structures

In this chapter, we will learn about data structures that will greatly aid our data science workflow.

## Vectors

Vectors are a sequence of values with the **same type**. We can create vectors using `c()`, which stands for "combine".

```{r vec}
(my_nums <- c(2.8, 3.2, 1.5, 3.8))
```

To access the elements inside a vector, we can do something called "slicing". To access a single item or multiple items, use the square bracket operator `[]`. In general `[]` in R means, "give me a piece of something". For example:

```{r basic-subsetting}
my_nums[4]
my_nums[1:3]
my_nums[c(1, 2, 3)] == my_nums[1:3]
```

In `my_nums[1:3]`, the `1:3` creates a vector from 1 to 3, which is then used to subset multiple items in a vector. Here are some additional useful functions:

```{r useful-fns, eval=F}
length(my_nums)
mean(my_nums)
max(my_nums)
min(my_nums)
sum(my_nums)
```

Given the data in the interactive block, consider the following exercises:

1. Select "Pouria" and "Ana" from the `names` vector.
1. Select all individuals who have ages greater than 20. Assume the order of names and ages correlates by index.
1. Select all individuals whose age is not 21.
1. Find the average age of all individuals.

<!---LEARNR EX 1-->

<iframe class="interactive" id="ch2-ex1" src="https://tig3r66.shinyapps.io/ch2-ex1/" scrolling="no" frameborder="no"></iframe>

<!------------->


### Missing values

So far we've worked with data with no missing values. In real life, however, we often have missing values (`NA` values). Unfortunately for us, R does not get along with `NA` values.

```{r missing-vals}
density_ha <- c(2.8, 3.2, 1.5, NA)
mean(density_ha)
```

Why did we get `NA`? Well, it's hard to say what a calculation including `NA` should be, so most calculations return `NA` when `NA` is in the data. One way to resolve this issue is to tell our function to remove the `NA` before executing:

```{r missing-vals-narm}
mean(density_ha, na.rm = TRUE)
```


## Lists

Lists are a vector-like structure that can store other objects/data structures. You can think of it like a vector that holds other vectors.

```{r list-intro}
sites <- c("a", "b", "c")
notes <- "It was a good day in the field today. Warm, sunny, lots of gators."
helpers <- 4
field_notes <- list(sites, notes, helpers)
```

You can index lists in the following ways:

```{r list-index}
field_notes[1]
field_notes[[1]]
```

We can also give the values names and access them using the `$` symbol--which is the preferred method--or via `["variable_name"]` with subsetting. Try getting the `my_sets` vector from `field_notes`.

<!---LEARNR EX 2-->

<iframe class="interactive" id="ch2-ex2" src="https://tig3r66.shinyapps.io/ch2-ex2/" scrolling="no" frameborder="no"></iframe>

<!------------->


## Data frames

This is where things get really exciting! We will use these data structures extensively in the upcoming labs, so it's important to pay attention here.

A data frame is a list of equal length vectors grouped together. More importantly, a data frame can contain both categorical and numerical values, whereas a vector can only contain variables of the same type (i.e., all numerical, all categorical, etc.).

```{R}
sites <- c("a", "a", "b", "c")
area_ha <- c(1, 2, 3, 4)
density_ha <- c(2.8, 3.2, 1.5, NA)
# creating the data frame
surveys <- data.frame(sites, density_ha, area_ha)
surveys
```

Here are some useful commands to investigate a data frame:

  * `str()` returns the structure of a data frame.
  * `length()` returns the length of a data frame.
  * `ncol()` returns the number of columns of a data frame (same as `length()`)
  * `nrow()` returns the number of rows of a data frame.

```{r df-str}
str(surveys)
ncol(surveys)
nrow(surveys)
```

Subsetting data frames is extremely similar to that for vectors. This time, however, we need to consider both rows and columns. We can access a specific member like this: `my_data_frame[row, column]`. Try playing around with the code below :)

<!---LEARNR EX 3-->

<iframe class="interactive" id="ch2-ex3" src="https://tig3r66.shinyapps.io/ch2-ex3/" scrolling="no" frameborder="no"></iframe>

<!------------->


### External data

We can read in external data using the`read.csv()` function. The main argument is the location of the data, which is either a url or a path on your computer.

```{r}
shrub_data <- read.csv('https://datacarpentry.org/semester-biology/data/shrub-dimensions-labeled.csv')
```

### Factors

Let's use the `str()` function to get more information about our variable `shrub_data`.

```{r}
str(shrub_data)
```

Notice that the `shrubID` column has type `Factor`. A factor is a special data type in R for categorical data. Factors are useful for statistics, but can mess up some aspects of computation as we'll see in future chapters.

```{r}
shrub_data <- read.csv('https://datacarpentry.org/semester-biology/data/shrub-dimensions-labeled.csv', stringsAsFactors = FALSE)
str(shrub_data)
```

<script>
  iFrameResize({}, ".interactive");
</script>
