# (PART) Data Wrangling {-}

# Data frames

Whenever you have rectangular, "spreadsheetey" data, your default data structure in R should be a data frame. Do not depart from this without good reason. Data frames are awesome because...

  * They neatly package related variables and maintain a "row-ordering" like that in a spreadsheet. This makes it easy to apply filters to rows and columns of interest.
  * Most functions for inference, modelling, and graphing are happy to be passed a data frame via a `data =` argument.
  * The set of packages known as the [`tidyverse`](https://www.tidyverse.org/) takes this one step further and explicitly prioritizes the processing of data frames. This includes popular packages like `dplyr` and `ggplot2`. In fact the tidyverse prioritizes a special flavor of data frame, called a "tibble."

Data frames -- unlike general arrays or matrices in R -- can hold variables of different flavours. For example, data frames can simultaneously hold *character data* (subject ID or name), *quantitative data* (white blood cell count), and *categorical information* (treated vs. untreated).

If you use data structures that can only hold 1 type of data (e.g., matrices) for data analysis, you are likely to make the terrible mistake of spreading a dataset out over multiple, unlinked objects. Why? Because you can't put character data, such as subject name, into the numeric matrix that holds white blood cell count. This fragmentation is a bad idea.


## Get the gapminder data

We will work with some of the data from the [Gapminder project](https://www.gapminder.org/). The Gapminder project contains the gapminder dataset, which summarises the progression of countries over time, looking at statistics like life expectancy and GDP. Let's first install the package:

```{r eval = FALSE}
install.packages("gapminder")
```

Now load the package:

```{r}
library(gapminder)
```


## Explore gapminder

By loading the `gapminder` package, we now have access to a data frame by the same name. Get an overview of this with `str()`, which displays the structure of an object. An object is essentially a storage unit for data.

```{r}
str(gapminder)
```

`str()` will provide a sensible description of almost anything and, worst case, nothing bad can actually happen. When in doubt, just use `str()` on some of the recently created objects to get some ideas about what to do next.

We could print the `gapminder` object itself to screen. However, if you've used R before, you might be reluctant to do this, because large datasets fill your Console and provide very little insight.

If you have not already done so, install the `tidyverse` meta-package now:

```{r eval = FALSE}
install.packages("tidyverse")
```

Now load it:

```{r, message = F, warning = F}
library(tidyverse)
```

Now we can print `gapminder` to screen! It is a tibble (and also a regular data frame) and the `tidyverse` provides a nice print method that shows the most important stuff and doesn't fill up your Console. The [tidyverse](https://github.com/hadley/tidyverse) offers a special case of R's default data frame: the `tibble`.

```{r}
# this is a comment
# anything that follows the pound sign (#) is not run
class(gapminder)  # the tibble is also a data frame
gapminder
```

Wow, that's a lot of output! The first line refers to what we're printing -- a tibble with 1704 rows and 6 columns. Below each column heading, you can see `<fct>   <fct>   <int>   <dbl>   <int>   <dbl>`. These refer to the **variable type**. That is, `fct` is a factor (kinda like a categorical variable), `int` stands for integer, and `dbl` stands for double (a number with decimal places).

If you are just dealing with data frames, you can rein in data frame printing explicitly with `head()` and `tail()`. `head()` displays the first 6 rows of your data frame by default whereas `tail()` shows the last 6 rows.

```{r}
head(gapminder)
tail(gapminder)
```

You can also specify the number of rows displayed by passing in a number.

```{r, eval}
head(gapminder, 10)
```

Just for your reference, if you want to change a data frame into a tibble for nicer printing, use `as_tibble()`!

```{r, eval = FALSE}
as_tibble(my_data_frame)  # my_data_frame is the thing we want to make a tibble
```

Here are more ways to query basic info on a data frame:

```{r}
names(gapminder)  # returns column names
ncol(gapminder)   # returns number of columns
nrow(gapminder)   # returns number of rows
dim(gapminder)    # returns # of rows x # of columns 
```

A statistical summary of the data can be obtained with `summary()`. That is, each column's statistics are shown separately.

```{r}
summary(gapminder)
```


### Importing and exporting data

We can export the data frame to a comma-separated values (.csv) file.

```{r}
write.csv(gapminder, file = "data/03_data-frames/gapminder.csv", sep = ",")
```

`.csv` stands for comma-separated values. This is the preferred way of importing and exporting data as it contains no formatting. This means the data can be easily parsed using various programming languages. You can also import a .csv file to Excel.

On top of writing to a .csv file, we can also read .csv files into R. It is super easy to do:

```{r, eval = FALSE}
gapminder2 <- read.csv("data/03_data-frames/gapminder.csv", header = TRUE)
as_tibble(gapminder2)  # converting the data frame to a tibble
```

As you can see,`read.csv()` returns a data frame by default.

Let's go back to the result of `str()` to talk about data frames.

```{r}
str(gapminder)
```

A data frame is a special case of a `list`, which is used in R to hold just about anything. Data frames are a special case where the length of each list component is the same. Data frames are (usually) superior to matrices in R because they can hold vectors of different flavors, e.g. numeric, character, and categorical data can be stored together.

### Exploring variables in a data Frame

**Note:** don't worry too much about the code to make the figures right now -- we will learn the specifics of data visualization in future lectures. In particular, we will introduce how to use `ggplot2`, a powerful package for data visualization that relies heavily on data frames. For now, focus on how the visualizations help to tell a story.

To specify a single variable from a data frame, use the dollar sign `$`. Let's explore the numeric variable for life expectancy.

```{r histogram-lifeExp}
head(gapminder$lifeExp)
summary(gapminder$lifeExp)
hist(gapminder$lifeExp)
```

You can also add custom labels to your graph:

```{r}
hist(gapminder$lifeExp,
     main="Histogram of Life Expectancy",
     xlab="Life expectancy (years)")
```

The `year` variable is an integer variable, but since there are so few unique values it also functions a bit like a categorical variable.

```{r}
summary(gapminder$year)
table(gapminder$year)
```

The variables for country and continent hold truly categorical information, which are stored as a `factor` in R.

```{r}
class(gapminder$continent)
summary(gapminder$continent)
levels(gapminder$continent)
nlevels(gapminder$continent)
```

The **levels** of the factor `continent` are "Africa", "Americas", etc. and this is what's usually presented to your eyeballs by R. In general, the levels are friendly human-readable character strings, like "male/female" and "control/treated". But *never ever ever* forget that, under the hood, R is really storing integer codes 1, 2, 3, etc. Look at the result from `str(gapminder$continent)` if you are skeptical.

```{r}
str(gapminder$continent)
```

Specifically in modelling and figure-making, factors are anticipated and accommodated by the functions and packages you will want to exploit. Note that factors are NOT integers. Factors are a numerical way that R uses to represent **categorical variables**.

### Plotting variables with R (OPTIONAL)

In this next bit, we count how many observations are associated with each continent and, as usual, try to portray that info visually. This makes it much easier to quickly see that African countries are well represented in this dataset.

```{r tabulate-continent}
table(gapminder$continent)
barplot(table(gapminder$continent))
```

Although we haven't begun our formal coverage of data visualization, it's so important for exploring a dataset that we will make a few figures anyway. Here we use only base R graphics, which are very basic.

When you are first presented with a data set and you have no idea what the relationships are between variables, a dot plot is usually a good starting point.

__Note:__ the squiggly line is called a tilde (~). In terms of the `plot()` function, the first variable is the dependent variable *y* and the second is/are the independent variable(s) *x*. That is, `plot(y ~ x, dataframe_name)`.

```{r first-plots-base-R}
plot(lifeExp ~ year, gapminder)
plot(lifeExp ~ gdpPercap, gapminder)
plot(lifeExp ~ log(gdpPercap), gapminder)
```

Important note: `log()` in R has base e. To use base 10, use the `log10()` function. Don't worry too much about logarithms yet :)


### Data frame summary

Use data frames and the [`tidyverse`](https://www.tidyverse.org/)!!! The `tidyverse` provides a special type of data frame called a "tibble" that has nice default printing behavior, among other benefits.

  * When in doubt, `str()` something or print something.
  * Always understand the basic extent of your data frames: number of rows and columns.
  * Understand what flavor the variables are.
  * Use factors!!! But with intention and care.
  * Do basic statistical and visual sanity checking of each variable.
  * Refer to variables by name (ex: `gapminder$lifeExp`) not by column number. Your code will be more robust and readable.


## Data frames with dplyr

`dplyr` is a package for data manipulation developed by Hadley Wickham and Romain Francois. It is built to be fast, highly expressive, and open-minded about how your data is stored. It is installed as part of the the [`tidyverse`](https://github.com/hadley/tidyverse) meta-package and it is among the packages loaded via `library(tidyverse)`.

Here's a fun bit of trivia: `dplyr` stands for "**d**ata frame **p**liers".


#### Load `dplyr` and `gapminder`

I choose to load the `tidyverse`, which loads `dplyr` among other packages we use incidentally below. Also load `gapminder`. If you don't have these packages, use the `install.packages('pkg_name')` command. Don't worry too much if you have warning messages.

```{r, message = F, warning = F}
library(tidyverse)
library(gapminder)
```

The `gapminder` data frame is a special kind of data frame: a tibble. You can treat tibbles just like a data frame.

```{r}
gapminder
```

It's tibble-ness is why we get nice compact printing. For a reminder of the problems with base data frame printing, go type `iris` in the R Console (`iris` is a well-known data frame built into R).

The `class()` function tells you what type of data structure you're dealing with. Note how gapminder's `class()` includes `tbl_df`, `tbl`, and `data.frame`. This shows that tibbles are built on top of the data frame.

```{r}
class(gapminder)
```

There will be some functions, like `print()`, that know about tibbles and do something special. There will others that do not, like `summary()`. In which case the regular data frame treatment will happen, because every tibble is also a regular data frame. As a reminder, `print()` prints stuff to the console and `summary()` provides a statistical overview of your dataset.

To turn any data frame into a tibble use `as_tibble()`:

```{r}
as_tibble(iris)
```

### Subsetting the data

If you feel the urge to store a little snippet of your data:

```{r}
canada <- gapminder[241:252, ]
```

Stop and ask yourself ...

> Do I want to create a separate subset of my original data?

If YES, **use proper data aggregation techniques or don’t subset the data.** Alternatively, only subset the data as a temporary measure while you develop your elegant code for computing on or visualizing these subsets.

If NO, then don't subset!

Copies and excerpts of your data clutter your workspace, invite mistakes, and sow general confusion. Avoid whenever possible. Reality can also lie somewhere in between. You will find the workflows presented below can help you accomplish your goals with minimal creation of temporary, intermediate objects.

If you create a variable you don't want, you can always remove it using the `rm()` function.

```{r, eval = FALSE}
x <- 'thing to not keep'
print(x)
rm(x)
# print(x)  # gives an error because x is deleted
```

### Filter rows with `filter()`

`filter()` takes logical expressions and returns the rows for which all are `TRUE`. Use this when you want to subset observations based on values.

The first argument is the name of the data frame. The subsequent arguments are the expressions that filter the dataframe. For example, let's filter all rows from `gapminder` where life expectancy is less than 29 years.

```{r}
filter(gapminder, lifeExp < 29)
```

When you run this line of code, `dplyr` filters the data and returns a new data frame. `dplyr` functions never modify their inputs, so if you want to save the result, you need to use the assignment operator, `<-`. Let's try this out! Here we filter based on `country` and `year`:

```{r}
rwanda_gthan_1979 <- filter(gapminder, country == "Rwanda", year > 1979)
```

Compare with some base R code to accomplish the same things:

```{r eval = FALSE}
gapminder[gapminder$lifeExp < 29, ]     # indexing is distracting
subset(gapminder, country == "Rwanda")  # almost same as filter; quite nice actually
```

What if you want to filter rows based on multiple values in a variable? For example, what if we want to filter all rows with either Rwanda or Afghanistan as countries?

```{r eval = FALSE}
filter(gapminder, country == "Rwanda" | country == "Afghanistan")
```

Here we use a Boolean operator, `|`, which means "or". Boolean operators always return either `TRUE` or `FALSE`. Some other common ones are `&` (and) and `!` (not).

What if we want to keep more than just 2 countries? One way would be to string Boolean operators together like so: `country == "Canada" | country == "Rwanda" | country == "Afghanistan | ...` This, however, is very wordy. A useful shortcut is to use `x %in% y`. This selects every row where `x` is one of the values in `y`:

```{r, eval = FALSE}
filter(gapminder, country %in% c("Rwanda", "Afghanistan"))
filter(gapminder, country %in% c("Canada", "Rwanda", "Afghanistan"))
```

Under no circumstances should you subset your data the way I did at first:

```{r eval = FALSE}
excerpt <- gapminder[241:252, ]
```

Why is this a terrible idea?

  * It is not self-documenting. What is so special about rows 241 through 252?
  * It is fragile. This line of code will produce different results if someone changes the row order of `gapminder`, e.g. sorts the data earlier in the script.
  
```{r eval = FALSE}
filter(gapminder, country == "Canada")
```

This call explains itself and is fairly robust.

### Pipe operator `%>%`

Before we go any further, we should exploit the new pipe operator that the tidyverse imports from the [`magrittr`](https://github.com/smbache/magrittr) package by Stefan Bache. Here's what it looks like: `%>%`. The RStudio keyboard shortcut: Ctrl + Shift + M (Windows), Cmd + Shift + M (Mac).

Let's demo then I'll explain:

```{r}
gapminder %>% head()
```

The above code is equivalent to `head(gapminder)`. The pipe operator takes the thing on the left-hand-side and __pipes__ it into the function call on the right-hand-side. It literally drops it in as the first argument. You can think of an argument as your input to a function. If you remember your grade school math, functions in R do exactly what you've learned in school -- it takes inputs (arguments/parameters) and spits an output, or a return value.

Never fear, you can still specify other arguments to this function! To see the first 3 rows of Gapminder, we could say `head(gapminder, 3)` or this:

```{r}
gapminder %>% head(3)
```

You are probably not impressed yet, but the magic will happen soon.

### Select Columns with `select()`

Use `select()` to subset the data on variables or columns. Here's a conventional call:

```{r}
select(gapminder, year, lifeExp)
```

And here's the same operation, but written with the pipe operator and piped through `head()`:

```{r}
gapminder %>%
  select(year, lifeExp) %>%
  head(4)
```

Think: "Take `gapminder`, then select the variables year and lifeExp, then show the first 4 rows."

If we didn't have the pipe operator, this is what the above function would look like:

```{r}
head(select(gapminder, year, lifeExp), 4)
```

As you can see, this is way harder to read. That's why the pipe operator is so useful.

An important note is that select does not actually filter any rows. It simply selects columns.

`select()` used alongisde `everything()` is also quite handy if you want to move variables within your data frame. The `everything()` function selects all variables not explicitly mentioned in `select()`. For example, let's move `year` and `continent` to the front of the `gapminder` tibble:

```{r}
select(gapminder, year, continent, everything())
```


Here's the data for Cambodia, but only certain variables...

```{r}
gapminder %>%
  filter(country == "Cambodia") %>%
  select(year, lifeExp)
```

... and what a typical base R call would look like:

```{r}
gapminder[gapminder$country == "Cambodia", c("year", "lifeExp")]
```

### Pure, predictable, pipeable (OPTIONAL)

We've barely scratched the surface of `dplyr` but I want to point out key principles you may start to appreciate. If you're new to R or "programming with data", feel free skip this section and [move on](block010_dplyr-end-single-table.html).

`dplyr`'s verbs, such as `filter()` and `select()`, are what's called [pure functions](http://en.wikipedia.org/wiki/Pure_function). To quote from Wickham's [Advanced R Programming book](http://adv-r.had.co.nz/Functions.html):

> The functions that are the easiest to understand and reason about are pure functions: functions that always map the same input to the same output and have no other impact on the workspace. In other words, pure functions have no side effects: they don’t affect the state of the world in any way apart from the value they return.

And finally, the data is __always__ the very first argument of the verb functions.


### Additional resources

`dplyr` official stuff

  * package home [on CRAN](http://cran.r-project.org/web/packages/dplyr/index.html)
    - note there are several vignettes, with the [introduction](http://cran.r-project.org/web/packages/dplyr/vignettes/introduction.html) being the most relevant right now
    - the [one on window functions](http://cran.rstudio.com/web/packages/dplyr/vignettes/window-functions.html) will also be interesting to you now
  * development home [on GitHub](https://github.com/hadley/dplyr)
  * [tutorial HW delivered](https://www.dropbox.com/sh/i8qnluwmuieicxc/AAAgt9tIKoIm7WZKIyK25lh6a) (note this links to a DropBox folder) at useR! 2014 conference

[RStudio Data Wrangling cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf), covering `dplyr` and `tidyr`. Remember you can get to these via *Help > Cheatsheets.* 

[Excellent slides](https://github.com/tjmahr/MadR_Pipelines) on pipelines and `dplyr` by TJ Mahr, talk given to the Madison R Users Group.

Blog post [Hands-on dplyr tutorial for faster data manipulation in R](http://www.dataschool.io/dplyr-tutorial-for-faster-data-manipulation-in-r/) by Data School, that includes a link to an R Markdown document and links to videos

[Cheatsheet](https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) from R Studio for `dplyr`.
