# Loops and Functions

Loops are fundamental a programming concept as they get a lot of repetitive stuff done in very few lines of code. Paired with custom functions, we can begin to tackle complex programming problems.


## For loops

Here's what the syntax of a `for` loop looks like:

```r
for (item in list_of_items) {
  do_something(item)
}
```

And here is an example:

```{r short-loop}
for (i in 1:3) {
  print(i)
}
```

In the previous example, we used the dummy variable `i` to take on some range of values. Notice that `i` can be called anything you want.

Try creating a for loop that prints the square of a number plus one for numbers ranging from 2 to 6.

<!---LEARNR EX 1-->

<iframe class="interactive" id="ch3-ex1" src="https://y-textbook.shinyapps.io/ch3-ex1/" scrolling="no" frameborder="no"></iframe>

<!------------->


### Looping over multiple files

We turn our attention now to a (slightly more) useful example: how do we analyze multiple files with similar contents?

In this hypothetical example, we have 5 datasets with satellite coodinates at specific points orbiting the Earth. Suppose the files are similarly named (click on the files to download them):

  * [locations-2016-01-01.txt](./data/04_intro-to-r/locations-2016-01-01.txt)
  * [locations-2016-01-02.txt](./data/04_intro-to-r/locations-2016-01-02.txt)
  * [locations-2016-01-03.txt](./data/04_intro-to-r/locations-2016-01-03.txt)
  * [locations-2016-01-04.txt](./data/04_intro-to-r/locations-2016-01-04.txt)
  * [locations-2016-01-05.txt](./data/04_intro-to-r/locations-2016-01-05.txt)

Our goal is to determine the number of satellite coordinates per file.

First, retrieve the name of each file.

```{r external-files}
my_dir <- "data/04_intro-to-r/"  # files are located in this location (on my computer)
my_files <- "locations-.*.txt"   # file names follow this pattern
data_files <- list.files(path = my_dir, pattern = my_files, full.names = TRUE)
```

Note that the asterisk in `"*.txt"` refers to "any name in this directory" whereas the `".txt"` part ensures we are only selecting .txt files.

Next, determine the number of observations in each file. We will assume that each row corresponds to a single coordinate.

```{r get-count}
results <- vector(mode = "integer", length = length(data_files))
for (i in 1:length(data_files)) {
  data <- read.csv(data_files[i])
  count <- nrow(data)
  results[i] <- count
}
```

Now, store the output in a data frame and associate the file name with the count.

```{r show-count-results}
# initializing the data frame with empty columns
results <- data.frame(file_name = character(length(data_files)),
                      count = integer(length(data_files)),
                      stringsAsFactors = FALSE)

# reading the data into the data frame
for (i in 1:length(data_files)) {
  data <- read.csv(data_files[i])
  count <- nrow(data)
  results$file_name[i] <- data_files[i]
  results$count[i] <- count
}

# voila!
results
```


### Nested loops

Sometimes, we need to loop over more than a single range of numbers. For example, what if we want to select all pixels on a 2x3 rectangular screen? Here, we need to cover both the "x" and "y" pixel coodinates:

```{r nested-loop}
for (i in 1:2) {
  for (j in 1:3) {
    print(paste("i = " , i, "; j = ", j, sep=""))
  }
}
```


## Functions

Sometimes, we will need to create custom functions. Luckily, we can define our own functions!

This is the general syntax for a function:
  
```{r gen-fn eval=F}
function_name <- function(arguments) {
  output_value <- do_something(inputs)
  return(output_value)
}
```

Remark: every function *returns* a value. Recall from your grade-school math class that functions take an input and return an output. In R, however, a function may or may not take user-defined input. This brings us to an extremely important point: **creating a function does NOT run it. You must call the function to run it.**

As an exercise, create a function called `calc_vol` that takes three parameters length, width, and height, and use those values to calculate the volume of the object. Then, call the function to calculate the volume of a 1x1x1 object and a 3x2x5 object.

<!---LEARNR EX 2-->

<iframe class="interactive" id="ch3-ex2" src="https://y-textbook.shinyapps.io/ch3-ex2/" scrolling="no" frameborder="no"></iframe>

<!------------->

Since R treats functions like a black box, you can't access a variable that was created in a function. You **must** save the output of a function (to a variable) to use it later.


### Conditionals within functions

We can use conditionals in a function for more complex tasks. As an exercise, create a function called `pred_c19_cases` to predict the number of COVID-19 cases in a population (note that these numbers are fictional):

1. The function will have two parameters `pop_size` (population size) and `vac_brand` (vaccine brand).
1. If the vaccine is Moderna, multiply `pop_size` by 0.941.
1. If the vaccine is Pfizer, multiply `pop_size` by 0.950.
1. If the vaccine is Astrazeneca, multiply `pop_size` by 0.870.
1. Return the predicted cases by subtracting the number of healthy individuals from `pop_size`.

<!---LEARNR EX 3-->

<iframe class="interactive" id="ch3-ex3" src="https://y-textbook.shinyapps.io/ch3-ex3/" scrolling="no" frameborder="no"></iframe>

<!------------->

Now that we've got the basics of R under our belts, we can jump into the delightful world of data science `r emo::ji("smile")`.

<script>
  iFrameResize({}, ".interactive");
</script>
