# (PART) Data Wrangling {-}

# Data exploration

Whenever you have rectangular, "spreadsheetey" data, your default data structure in R should be the **data frame**. Data frames are awesome because

  1. They neatly package related variables and maintain a "row-ordering" like that in a spreadsheet. This makes it easy to apply filters to rows and columns of interest.
  2. Most functions for inference, modelling, and graphing will happily take a data frame object.
  3. The set of packages known as the [tidyverse](https://www.tidyverse.org/) takes data frames one step further and explicitly prioritizes the processing of data frames.

Data frames, unlike general arrays or matrices in R, can hold variables of different flavours. For example, data frames can simultaneously hold *character data* (e.g., subject ID or name), *quantitative data* (e.g., white blood cell count), and *categorical information* (e.g., treated vs. untreated).

If you use data structures that can only hold 1 type of data (e.g., matrices) for data analysis, you are likely to make the terrible mistake of spreading a dataset out over multiple, unlinked objects. Why? Because you can't put character data, such as subject name, into the numeric matrix that holds white blood cell count.


## Get the gapminder data

We will work with some of the data from the [Gapminder project](https://www.gapminder.org/). The Gapminder project contains the gapminder dataset, which summarises the progression of countries over time for statistics like life expectancy and GDP.

If you haven't installed gapminder or the tidyverse yet, you can do so like this:

```{r eval = FALSE}
install.packages("gapminder")
install.packages("tidyverse")
```

Now load the two packages.

```{r loading, message=FALSE, warning=FALSE}
library(gapminder)
library(tidyverse)
```


## Explore gapminder

By loading the gapminder package, we now have access to a data frame by the same name.

```{r gap-view}
gapminder
class(gapminder)
```

Notice that the **class** (type of data structure) of the gapminder object is a **tibble**, the tidyverse's version of R's data frame. A tibble is also a data frame.

Although this seems like a lot of output, notice that tibbles provide a nice print method that shows the most important stuff and doesn't fill up your console. Let's make sense of the output:

  1. The first line refers to what we're printing---a tibble with 1704 rows and 6 columns.
  1. Below each column heading, we see `<fct>   <fct>   <int>   <dbl>   <int>   <dbl>`. These refer to the **variable type** of that column.
      a. `fct` is a factor (kind of like a categorical variable), `int` stands for integer, and `dbl` stands for double (a number with decimal places).

If you're only interested in a summary, use `str()`, `head()` or `tail()`.
  
  1. `str()` will provide a sensible description of almost anything and, worst case, nothing bad can actually happen. When in doubt, just use `str()` on your recently created objects to get ideas about what to do next.  
  1. `head()` displays the first 6 rows of your data frame by default, and
  1. `tail()` shows the last 6 rows.

Play around with these functions in the interactive block below!

<!---LEARNR EX 1-->

<iframe class="interactive" id="ch4-ex1" src="https://y-textbook.shinyapps.io/ch4-ex1/" scrolling="no" frameborder="no"></iframe>

<!------------->

Just for your reference, if you want to change a data frame into a tibble for nicer printing, use `as_tibble()`!

```{r, eval = FALSE}
as_tibble(my_data_frame)  # my_data_frame is the thing we want to make a tibble
```

Here are more ways to query basic info on a data frame:

| Function   | Description                                  |
|:-----------|:---------------------------------------------|
| `names()`  | returns column names                         |
| `ncol()`   | returns number of columns                    |
| `nrow()`   | returns number of rows                       |
| `dim()`    | returns # of rows by # of columns            |
| `summary()`| returns a statistical summary of each column |

Try playing around with these functions in the interactive window.

<!---LEARNR EX 2-->

<iframe class="interactive" id="ch4-ex2" src="https://y-textbook.shinyapps.io/ch4-ex2/" scrolling="no" frameborder="no"></iframe>

<!------------->


### Importing and exporting data

We can export data frames to a comma-separated values (.csv) file.

```{r}
write.csv(gapminder, file = "data/03_data-frames/gapminder.csv")
```

The ".csv" extension stands for comma-separated values. This is the preferred way of importing and exporting data as it contains no formatting.

On top of writing to a .csv file, we can also read .csv files into R. It's as simple as `read.csv()`!

```{r}
gapminder2 <- read.csv("data/03_data-frames/gapminder.csv", header = TRUE)
class(gapminder2)
```

As you can see,`read.csv()` returns a data frame object by default.


### Exploring variables in a data frame

To specify a single variable from a data frame, use the dollar sign `$`. Let's explore the numeric variable for life expectancy.

```{r histogram-lifeExp}
head(gapminder$lifeExp)
summary(gapminder$lifeExp)
hist(gapminder$lifeExp)
```

Don't worry too much about the code to make the figures right now---we will learn how to visualize data in future lectures. For now, let's continue to explore gapminder.

Take a look at the `year` variable:

```{r}
class(gapminder$year)
```

Notice that `year` holds integers. On the other hand, `continent` holds categorical information, which is called a **factor** in R.

```{r}
class(gapminder$continent)
```

Now, I want to illustrate something important:

```{r}
summary(gapminder$year)
summary(gapminder$continent)
```

Notice that the same function returned different outputs for different variable types---forgetting this observation can lead to confusion in the future, so make sure to check your data before analysis! Let's check out a couple more useful functions and highlight important ideas in the meantime.

Within a given column/variable,

  1. `table()` returns the number of observations,
  2. `levels()` returns unique values, and
  3. `nlevels()` returns the number of unique values.

```{r}
table(gapminder$continent)
levels(gapminder$continent)
nlevels(gapminder$continent)
```

The **levels** of the factor `continent` are "Africa", "Americas", etc.---this is what's usually presented to your eyeballs by R. Behind the scenes, R assigns integer values (i.e., 1, 2, 3, ...) to each level. *Never ever ever* forget this fact. Look at the result from `str(gapminder$continent)` if you are skeptical:

```{r}
str(gapminder$continent)
```

Specifically in modelling and figure-making, factors are anticipated and accommodated by the functions and packages you will want to exploit. Note that factors do NOT contain integers. Factors are a numerical way that R uses to represent categorical data.

Tl;dr, factors are categorical variables whereas levels are unique values within a factor.


### Data frame summary

Use data frames and the [`tidyverse`](https://www.tidyverse.org/)!! The `tidyverse` provides a special type of data frame called a "tibble" that has nice default printing behavior, among other benefits.

  * When in doubt, `str()` something or print something.
  * Always understand the basic extent of your data frames: number of rows and columns.
  * Understand what your variable types are.
  * Use factors!! (but with intention and care)
  * Do basic statistical and visual sanity checking of each variable.
  * Refer to variables by name (ex: `gapminder$lifeExp`) and NOT by column number. Your code will be more robust and readable.


## Data frames with dplyr

`dplyr` is a package for data manipulation developed by Hadley Wickham and Romain Francois. It is built to be fast, highly expressive, and open-minded about how your data is stored. It is installed as part of the the [`tidyverse`](https://github.com/hadley/tidyverse) meta-package and it is among the packages loaded via `library(tidyverse)`.

Here's a bit of fun trivia: `dplyr` stands for "**d**ata frame **p**liers".


### Subsetting data

If you feel the urge to store a little snippet of your data:

```{r}
canada <- gapminder[241:252, ]
```

Stop and ask yourself ...

> Do I want to create a separate subset of my original data?

If "YES," **use proper data aggregation techniques or don’t subset the data.** Alternatively, only subset the data as a *temporary measure* while you develop your elegant code.

If "NO," then don't subset!

Copies and excerpts of your data clutter your workspace, invite mistakes, and sow general confusion. Avoid whenever possible. Reality can also lie somewhere in between. You will find the workflows presented below can help you accomplish your goals with minimal creation of temporary, intermediate objects.

Recall the`rm()` function, which removes unwanted variable(s).

```{r, eval = FALSE}
x <- 'thing to not keep'
print(x)
rm(x)
# print(x)  # gives an error because x is deleted
```


### Filter rows with `filter()`

`filter()` takes logical expressions and returns the rows for which all are `TRUE`. Use this when you want to subset observations based on values.

The first argument is the name of the data frame. The subsequent arguments are the expressions that filter the dataframe. For example, let's filter all rows from `gapminder` where life expectancy is less than 29 years.

```{r}
filter(gapminder, lifeExp < 29)
```

When you run this line of code, `dplyr` filters the data and returns a new data frame. `dplyr` functions never modify their inputs, so if you want to save the result, you need to use the assignment operator, `<-`. Let's try this out! Here we filter based on `country` and `year`:

```{r}
rwanda_gthan_1979 <- filter(gapminder, country == "Rwanda", year > 1979)
```

Compare with some base R code to accomplish the same things:

```{r eval = FALSE}
gapminder[gapminder$lifeExp < 29, ]     # indexing is distracting
subset(gapminder, country == "Rwanda")  # almost same as filter; quite nice actually
```

What if you want to filter rows based on multiple values in a variable? For example, what if we want to filter all rows with either Rwanda or Afghanistan as countries?

```{r eval = FALSE}
filter(gapminder, country == "Rwanda" | country == "Afghanistan")
```

Here we use a Boolean operator, `|`, which means "or". Boolean operators always return either `TRUE` or `FALSE`. Some other common ones are `&` (and) and `!` (not).

What if we want to keep more than just 2 countries? One way would be to string Boolean operators together like so: `country == "Canada" | country == "Rwanda" | country == "Afghanistan | ...` This, however, is very wordy. A useful shortcut is to use `x %in% y`. This selects every row where `x` is one of the values in `y`:

```{r, eval = FALSE}
filter(gapminder, country %in% c("Rwanda", "Afghanistan"))
filter(gapminder, country %in% c("Canada", "Rwanda", "Afghanistan"))
```

Under no circumstances should you subset your data the way I did at first:

```{r eval = FALSE}
excerpt <- gapminder[241:252, ]
```

Why is this a terrible idea?

  * It is not self-documenting. What is so special about rows 241 through 252?
  * It is fragile. This line of code will produce different results if someone changes the row order of `gapminder`, e.g. sorts the data earlier in the script.
  
```{r eval = FALSE}
filter(gapminder, country == "Canada")
```

This call explains itself and is fairly robust.

### Pipe operator `%>%`

Before we go any further, we should exploit the new pipe operator that the tidyverse imports from the [`magrittr`](https://github.com/smbache/magrittr) package by Stefan Bache. Here's what it looks like: `%>%`. The RStudio keyboard shortcut: Ctrl + Shift + M (Windows), Cmd + Shift + M (Mac).

Let's demo then I'll explain:

```{r}
gapminder %>% head()
```

The above code is equivalent to `head(gapminder)`. The pipe operator takes the thing on the left-hand-side and __pipes__ it into the function call on the right-hand-side. It literally drops it in as the first argument. You can think of an argument as your input to a function. If you remember your grade school math, functions in R do exactly what you've learned in school -- it takes inputs (arguments/parameters) and spits an output, or a return value.

Never fear, you can still specify other arguments to this function! To see the first 3 rows of Gapminder, we could say `head(gapminder, 3)` or this:

```{r}
gapminder %>% head(3)
```

You are probably not impressed yet, but the magic will happen soon.

### Select Columns with `select()`

Use `select()` to subset the data on variables or columns. Here's a conventional call:

```{r}
select(gapminder, year, lifeExp)
```

And here's the same operation, but written with the pipe operator and piped through `head()`:

```{r}
gapminder %>%
  select(year, lifeExp) %>%
  head(4)
```

Think: "Take `gapminder`, then select the variables year and lifeExp, then show the first 4 rows."

If we didn't have the pipe operator, this is what the above function would look like:

```{r}
head(select(gapminder, year, lifeExp), 4)
```

As you can see, this is way harder to read. That's why the pipe operator is so useful.

An important note is that select does not actually filter any rows. It simply selects columns.

`select()` used alongisde `everything()` is also quite handy if you want to move variables within your data frame. The `everything()` function selects all variables not explicitly mentioned in `select()`. For example, let's move `year` and `continent` to the front of the `gapminder` tibble:

```{r}
select(gapminder, year, continent, everything())
```


Here's the data for Cambodia, but only certain variables...

```{r}
gapminder %>%
  filter(country == "Cambodia") %>%
  select(year, lifeExp)
```

... and what a typical base R call would look like:

```{r}
gapminder[gapminder$country == "Cambodia", c("year", "lifeExp")]
```

### Pure, predictable, pipeable (OPTIONAL)

We've barely scratched the surface of `dplyr` but I want to point out key principles you may start to appreciate. If you're new to R or "programming with data", feel free skip this section.

`dplyr`'s verbs, such as `filter()` and `select()`, are what's called [pure functions](http://en.wikipedia.org/wiki/Pure_function). To quote from Wickham's [Advanced R Programming book](http://adv-r.had.co.nz/Functions.html):

> The functions that are the easiest to understand and reason about are pure functions: functions that always map the same input to the same output and have no other impact on the workspace. In other words, pure functions have no side effects: they don’t affect the state of the world in any way apart from the value they return.

And finally, the data is **always** the very first argument of every dplyr function.


### Additional resources

`dplyr` official stuff

  * package home [on CRAN](http://cran.r-project.org/web/packages/dplyr/index.html)
    - note there are several vignettes, with the [introduction](http://cran.r-project.org/web/packages/dplyr/vignettes/introduction.html) being the most relevant right now
    - the [one on window functions](http://cran.rstudio.com/web/packages/dplyr/vignettes/window-functions.html) will also be interesting to you now
  * development home [on GitHub](https://github.com/hadley/dplyr)
  * [tutorial HW delivered](https://www.dropbox.com/sh/i8qnluwmuieicxc/AAAgt9tIKoIm7WZKIyK25lh6a) (note this links to a DropBox folder) at useR! 2014 conference

[RStudio Data Wrangling cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf), covering `dplyr` and `tidyr`. Remember you can get to these via *Help > Cheatsheets.* 

[Excellent slides](https://github.com/tjmahr/MadR_Pipelines) on pipelines and `dplyr` by TJ Mahr, talk given to the Madison R Users Group.

Blog post [Hands-on dplyr tutorial for faster data manipulation in R](http://www.dataschool.io/dplyr-tutorial-for-faster-data-manipulation-in-r/) by Data School, that includes a link to an R Markdown document and links to videos

[Cheatsheet](https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) from R Studio for `dplyr`.

<script>
  iFrameResize({}, ".interactive");
</script>
