---
title: "Session 2 Review"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "This is a review of R basics."
---

```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.timelimit = 60)
knitr::opts_chunk$set(error = TRUE)
# loading packages
library(gapminder)
library(tidyverse)
```


## Welcome

In this brief lesson, we will explore a dataset called "gapminder" and review basic data wrangling techniques. These lessons correspond to chapter 4 of the [Youreka textbook](https://youreka-textbook.netlify.app/more-data-frames.html).

At this point in our lessons, I would like to take a minute to look at the bigger picture: the R programming skills you earn in our program are designed to be *tools* that you can use for your project. That is, R will be a "means to an end" for your projects. That being said, programming is super cool, so let's continue the fun!

## Exploring gapminder

### str(), head(), tail()

First, load the gapminder library then take a brief look at it. If you haven't installed the gapminder package, you can do so like this: `install.packages("gapminder", dependencies = TRUE)`

```{r load-gap}
library(gapminder)
gapminder
```

Although it's nice to see all the individual observations, sometimes we just want a brief look into the data. This is where `str()`, `head()`, and `tail()` come in handy. Try running the code below to see what happens!

```{r preview-str, exercise=TRUE, exercise.eval=F}
str(gapminder)
```

Recall that `head()` returns the first 6 rows of your data frame and `tail()` returns the last 6 rows of your data frame. You can even specifiy the number of entries you want by passing a number to `n = `. Try showing the first 7 entries of gapminder.

```{r preview-head, exercise=TRUE}

```
```{r preview-head-solution}
head(gapminder, n = 7)  # alternatively: head(gapminder, 7)
```

If you want to get really fancy, using `head()` on the "lifeExp" column in gapminder.

```{r head-var, exercise=TRUE}

```
```{r head-var-solution}
head(gapminder$lifeExp)
```

Now, I want you to explore the "continent" column in gapmidner using the `table()`, `levels()`, and `nlevels()` functions on your own:

```{r other-explore, exercise=TRUE}

```
```{r other-explore-solution}
table(gapminder$continent)   # returns number of observations
levels(gapminder$continent)  # returns unique values in a column
levels(gapminder$continent)  # returns number of unique values in a column
```


### Getting more information from data frames

Sometimes, we want to see the column names of a data frame to know what we're working with. We can do so using the `names()` function.

```{r names}
names(gapminder)
```

Notice that `names()` returns an array that you can save. As an exercise, try checking if the column names contains "year" and "popn".

```{r year-check, exercise=TRUE}

```
```{r year-check-solution}
c("year", "popn") %in% names(gapminder)
```

We can also get a statistical summary of a data frame using `summary()`.

```{r summary}
summary(gapminder)
```


## Subsetting data with dplyr

### Introduction to filter()

Dplyr will be your best friend during data wrangling. Here's a fun piece of trivia: d = data and p = pliers, so d + plier = data pliers :)

First, load dplyr by loading the tidyverse:

```{r load-dplyr, warning=FALSE, message=FALSE}
library(tidyverse)
```

Imagine that you want to find rows in a column that match a specific condition, say all rows in gapminder where "lifeExp" is less than 29. We can check this using `filter()`.

```{r filter-intro}
filter(gapminder, lifeExp < 29)
```

We can combine conditions using a comma, or Boolean operators:

```{r combine-intro}
filter(gapminder, country == "Rwanda", year > 1979)  # finding all rows where country is Rwanda and year > 1979
```

As an exercise, try executing the above code using the logical "and" `&`.

```{r logical-and, exercise=TRUE}

```
```{r logical-and-solution}
filter(gapminder, country == "Rwanda" & year > 1979)
```


## Filter exercises

Use the code chunks below to find all entries in gapminder that:

1. Have country as Rwanda or Afghanistan

  ```{r country-filter, exercise=TRUE}

  ```
  ```{r country-filter-solution}
    filter(gapminder, country == "Rwanda" | country == "Afghanistan")  # | is the logical "or"
  ```

2. Have country as Rwanda, Afghanistan, or Canada.

  ```{r multiple-vals, exercise=TRUE}

  ```
  ```{r multiple-vals-solution}
  # method 1
  filter(gapminder, country %in% c("Rwanda", "Afghanistan", "Canada"))
  # method 2
  filter(gapminder, country == "Rwanda" | country == "Afghanistan" | country == "Canada")
  ```

3. Have year equal to 1952 or 1962.

  ```{r year-filter, exercise=TRUE}
  
  ```
  ```{r year-filter-solution}
  # method 1
  filter(gapminder, year %in% c(1952, 1962))
  # method 2
  filter(gapminder, year == 1952 | year == 1962)
  ```

4. Have year as 1952 or 1962 AND where country is Canada.

  ```{r multi-condition-and, exercise=TRUE}
  
  ```
  ```{r multi-condition-and-solution}
  # method 1
  filter(gapminder, year %in% c(1952, 1962), country == "Canada")
  # method 2
  filter(gapminder, year %in% c(1952, 1962) & country == "Canada")
  ```

5. Have year as 1952 or 1962 OR where country is Canada.

  ```{r multi-condition-or, exercise=TRUE}

  ```
  ```{r multi-condition-or-solution}
  filter(gapminder, year %in% c(1952, 1962) | country == "Canada")
  ```


## Pipe operator %>%

Note: you need to load the tidyverse to get the pipe operator.

The pipe operator `%>%` allows us to clean up our code. Recall from the textbook that `%>%` takes whatever is on the left side and sticks it into the first function argument on the right side. For example:

```{r pipe-intro}
gapminder %>% head()
```

Now, try using the pipe operator to first filter gapminder where lifeExp is less than 29 and then view the first four columns.

```{r pipe-ex1, exercise=TRUE}

```
```{r pipe-ex1-solution}
gapminder %>%
  filter(lifeExp < 29) %>%
  head(4)

# note that R doesn't care about indentation levels
# try not to do this if you have a long chain of functions for code readability
gapminder %>% filter(lifeExp < 29) %>% head(4)
```

## select()

Imagine we have a data frame with lots of columns and we only want to select 2 for our analysis. Then, we can use the `select()` function from dplyr.

```{r select-intro}
gapminder %>% select(year, lifeExp)
# alternatively: select(gapminder, year, lifeExp)
```

`select()` can also be used to reorder columns. Another helpful function is `everything()`, which selects all columns in a data frame. For example, let's bring the "year" and "continent" columns in gapminder to the first entry.

```{r col-reorder}
gapminder %>% select(year, continent, everything())
```

Let's put what we've learned altogether now! First, filter gapminder where country is Canada and year is 1952. Second, select the "year", "country", and "lifeExp" columns. Third, view the first 10 rows.

```{r comprehensive-ex, exercise=TRUE}

```
```{r comprehensive-ex-solution}
gapminder %>%
	filter(country == "Canada" & year == 1952) %>%
	select(year, country, lifeExp) %>%
	head(10)
```



