---
title: "Session 4 Review"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "This is a review of R basics."
---

```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.timelimit = 60)
knitr::opts_chunk$set(error = TRUE)
# loading packages
library(gapminder)
library(tidyverse)
```

## Welcome

In this brief lesson, we will explore a dataset called "gapminder" and review basic data wrangling techniques. These lessons correspond to chapter 4 of the [Youreka textbook](https://youreka-textbook.netlify.app/more-data-frames.html).

At this point in our lessons, I would like to take a minute to look at the bigger picture: the R programming skills you earn in our program are designed to be *tools* that you can use for your project. That is, R will be a "means to an end" for your projects. That being said, programming is super cool, so let's continue the fun!

## Working with data frames

When we are working with R, most of the time we will find ourselves dealing with data frames. The previous two sessions have been dedicated to the `dplyr` package. In this session we will do more exercises within the r-base.

First of all, let's refresh ourselves with the gapminder dataset.

```{r load-gap}
head(gapminder)
```

What are the names of the columns of the data frame `gapminder`?

```{r names, exercise=TRUE}

```
```{r names-solution}
names(gapminder)
```

### Calling and subsetting data frames

Now, please subset a column called `lifExp` using `dplyr` pipeline.

```{r column-dplyr, exercise=TRUE}
gapminder %>%
```
```{r column-dplyr-solution}
gapminder %>% select(lifeExp)
```

Please call this column using r-base by the name or by the column number.

```{r column-base, exercise=TRUE}
gapminder$
gapminder[]
```
```{r column-base-solution}
gapminder$lifeExp
gapminder[[4]]
```
  
Please note that each column of the data frame is a vector.

Now I would like to subset the data frame with entries only in the 1960s and 1970s. Please try it with appropriate `dplyr` functions.

```{r row-dplyr, exercise=TRUE}
gapminder %>%
```
```{r row-dplyr-solution}
gapminder %>% filter(year >= 1960, year <= 1979)
```

Let's do the same thing using r-base. Hint: the entries in the 1960s occupy only certain rows. Therefore, you will want to determine which rows contain the data we want with logical operation and then call those rows.

```{r row-base, exercise=TRUE}
gapminder[ , ]
```
```{r row-base-solution}
gapminder[gapminder$year >= 1960 & gapminder$year <= 1979, ]
```

Please note that "&" stands for the logical AND operation. Only TRUE & TRUE will return TRUE. Also note that we did not write anything after the comma. This means we include all the columns, rather than no columns.

### Searching for patterns in strings

Next let's find the countries that contains "Congo" in their names. To search for a pattern in a string, we need to use the function `grepl`. This function takes at least 2 parameters, the first being the search pattern, the second being the vector in which it searches. Please try it within the `filter()` function of `dplyr`.

```{r row-grepl, exercise=TRUE}
gapminder %>% filter(grepl( ))
```
```{r row-grepl-solution}
gapminder %>% filter(grepl("Congo", country))
```

Similarly, you can also use it with the r-base.

```{r grepl-base, exercise=TRUE}
gapminder[ , ]

```
```{r grepl-base-solution}
gapminder[grepl("Congo", gapminder$country), ]
```

### Cautions for `levels`

How many countries do we have in the gapminder dataset with names containing "Congo"? Please try the following block.

```{r levels, exercise=TRUE, exercise.eval=F}
levels(gapminder[grepl("Congo", gapminder$country), ]$country)
```

Wow, this result certainly does not make sense. This is because, although we selected only a few rows, the column `country` could still accept many potential factor-type values, even though they are not present in the subset we are looking at. Please note `levels()` function may lead to such artefacts and needs extra caution.

`unique()` is one way to address this issue.

```{r unique, exercise=TRUE, exercise.eval=F}
unique(gapminder[grepl("Congo", gapminder$country), ]$country)
```

<!-- ### Putting all together -->

<!-- Now, please use the methods we've practiced above to make a dataset with entries from country names containing "Congo" in 1960s and 1970s with only the columns of country,year and lifeExp and gdpPercap. Save the result as a new data frame `df`. -->

<!-- ```{r complex-base, exercise=TRUE} -->

<!-- ``` -->
<!-- ```{r complex-base-solution} -->
<!-- df <- gapminder %>% -->
<!--   filter(grepl("Congo", country), -->
<!--          year >= 1960, -->
<!--          year <= 1970) %>% -->
<!--   select(country, year, lifeExp, gdpPercap) -->

<!-- df <- gapminder[grepl("Congo", gapminder$country) & -->
<!--             gapminder$year >= 1960 & -->
<!--             gapminder$year <= 1979, -->
<!--           c(1,3,4,5)] -->

<!-- df <- gapminder[grepl("Congo", gapminder$country) & -->
<!--             gapminder$year >= 1960 & -->
<!--             gapminder$year <= 1979, -->
<!--           c("country", "year", "lifeExp", "gdpPercap")] -->
<!-- ``` -->

<!-- ## Performing t-test -->

<!-- ### Two ways of writing the command -->

<!-- Is the life expectancy of the two countries with names containing "Congo" significantly different over 1960s and 1970s? To answer this question, we will need to use t-test. -->

<!-- First let's save their life expectancy into two vectors. Then we can pass these two vectors into `t.test()`. Is this a paired or unpaired test? -->

<!-- ```{r two-vec-t, exercise=TRUE} -->
<!-- DRCongo <- -->
<!-- ROCongo <- -->
<!-- t.test( , ) -->
<!-- ``` -->
<!-- ```{r two-vec-t-solution} -->
<!-- DRCongo <- df[df$country == "Congo, Dem. Rep.",]$lifeExp -->
<!-- ROCongo <- df[df$country == "Congo, Rep.",]$lifeExp -->
<!-- t.test(DRCongo, ROCongo, paired = TRUE) -->
<!-- ``` -->

<!-- Another way is to use the sign `~` between the response variable and the explanatory variable(s). -->

<!-- ```{r conn-t, exercise=TRUE} -->
<!-- t.test( ~ , ) -->
<!-- ``` -->
<!-- ```{r conn-t-solution} -->
<!-- t.test(lifeExp ~ country, data = df, paired = TRUE) -->
<!-- ``` -->

<!-- ### Paired t-test is actually a one-sample t-test -->

<!-- We have mentioned in the lecture that the paired t-test is actually a one-sample t-test. The differences of the paired data points are compared with 0. Could you perform this two-sample paired t-test in the one-sample t-test manner? -->

<!-- ```{r alt-paired, exercise=TRUE} -->
<!-- t.test( ~ , ) -->
<!-- ``` -->
<!-- ```{r alt-paired-solution} -->
<!-- t.test(DRCongo - ROCongo, mu=0) -->
<!-- ``` -->

<!-- Please compare the t, df and p-value with the results above. What do you find? -->

<!-- ### Performing t-test in batch -->

<!-- Can you compare the life expectancy of countries in Europe and Americas in 1952? Is this a paired or unpaired t-test? -->

<!-- Let's create a truncated data frame with entries from 1952 and the two continents first. Then we use the wave sign for the t-test. -->

<!-- ```{r unpaired, exercise=TRUE, exercise.eval=F} -->
<!-- temp.df <- -->
<!-- t.test( ~ , )) -->
<!-- ``` -->
<!-- ```{r unpaired-solution} -->
<!-- temp.df <- gapminder %>% -->
<!--   filter(continent %in% c("Europe", "Americas"), year == 1952) -->
<!-- t.test(lifeExp ~ continent, data = temp.df) -->
<!-- ``` -->

<!-- we can also return the p-value and mean of difference -->

<!-- ```{r unpaired-return, exercise=TRUE, exercise.eval=F} -->
<!-- test <- t.test(lifeExp ~ continent, data = temp.df) -->
<!-- test$p.value -->
<!-- test$estimate -->
<!-- test$estimate[1] -->
<!-- test$estimate[2] -->
<!-- ``` -->

<!-- We would like to calculate such difference for every year, and save the results in a new data frame. -->

<!-- ```{r empty, exercise=TRUE, exercise.eval=F} -->
<!-- results.df <- data.frame() -->
<!-- results.df <- rbind(results.df, -->
<!--                     c(1952, test$p.value, test$estimate)) -->
<!-- names(results.df) <- c("year", -->
<!--                          "p.value", -->
<!--                          "mean.Americas", -->
<!--                          "mean.Europe") -->
<!-- ``` -->

<!-- Could you use a for loop to complete this task? -->

<!-- ```{r for, exercise=TRUE, exercise.eval=F} -->
<!-- for (  ) { -->

<!-- } -->
<!-- ``` -->
<!-- ```{r for-solution} -->
<!-- for (i in levels(as.factor(gapminder$year))[-1]) { -->
<!--   temp.df <- gapminder %>% -->
<!--   filter(continent %in% c("Europe", "Americas"), -->
<!--          year == i) -->
<!--   test <- t.test(lifeExp ~ continent, data = temp.df) -->
<!--   results.df <- rbind(results.df, -->
<!--                     c(i, test$p.value, test$estimate)) -->
<!-- } -->
<!-- results.df -->
<!-- ``` -->

<!-- Now add a new column called "difference" as the difference between two mean values. -->

<!-- ```{r difference, exercise=TRUE, exercise.eval=F} -->
<!-- results.df <- data.frame(results.df, -->
<!--                          difference = as.numeric(results.df$mean.Europe) - as.numeric(results.df$mean.Americas)) -->
<!-- results.df -->
<!-- ``` -->

<!-- Look at the data frame. What conclusions can you draw? -->
