---
title: "Session 1 Review"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "This is a review of R basics."
---

```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.timelimit = 60)
knitr::opts_chunk$set(error = TRUE)
# global variables
shoe_price <- 25
vac_price <- c(5, 10, 3, 25)
```


## Welcome

In this brief lesson, you will review conditionals, vectors, and data frames. These lessons correspond to chapters 2-3.1 of the [Youreka textbook](https://youreka-textbook.netlify.app/basics-of-r-part-2.html).

If you are ready to begin, click on!

## Conditionals

### Logical comparisons

We can use logical operators to compare objects in R. For example, compare if a vaccine price is less than $3.

```{r gthan, exercise=TRUE}
vaccine_price <- 5
# compare here

```
```{r gthan-solution}
vaccine_price > 3
```


### If, else if, and else statements

Recall that `if`, `else if`, and `else` statements help control the flow of your program. For example, we can check if the price of a shoe is less than $50 (and if it is, we buy it!).

```{r shoe-purchase, exercise=T, exercise.eval=F}
shoe_price <- 25
if (shoe_price < 50) {
  print("Let's buy this shoe!")
} else {
  print("Maybe we'll wait until there's a sale...")
}
```


## Vectors

Vectors were the first data structure that we learned. Try creating a vector with the following vaccine prices: 5, 10, 3, 25.

```{r vac-price, exercise=TRUE}
vac_price <-
```
```{r vac-price-solution}
vac_price <- c(5, 10, 3, 25)
```

Now, find the average price of the vaccine price vector.
```{r mean-price, exercise=TRUE}

```
```{r mean-price-solution}
mean(vac_price)
```

Sometimes, our data has missing values. Try finding the mean of the following vaccine prices without any NA values.
```{r na-values, exercise=TRUE}
vac_price <- c(4, NA, 2, 15, 4)
# solve here

```
```{r na-values-solution}
mean(vac_price, na.rm=TRUE)
```

Note that data frames can only include objects of the **same** type. Look at what happens if we include an integer with character data.
```{r same-type}
c("hello", "goodbye", 0)
```

Notice that R *casts* `0` to the character `"0"`.

Let's review data frames now!


## Data Frames

Given the following vectors, create a data frame called "surveys".

```{r create-df, exercise=TRUE}
provinces <- c("AB", "BC", "ON", "QB")
soil_density <- c(3, 5, 1.5, NA)
area <- c(3, 5, 1.9, 2.7)
# create the data frame here
```
```{r create-df-solution}
surveys <- data.frame(provinces, soil_density, area)
```








